Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-03-02 src/tex/ms.tex
Update ms.tex
merged in most of Will's comments

We carry out \ac{PE} using a custom version of \textsc{Bilby} \citep{Bilby},
that applies Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline
waveform. We take the \ac{PE} configuration as in \citet{GWTC-2.1, GWTC-3,
GWTC-2.1_dataset, GWTC-3_dataset} as a starting point, with
\textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb} as the reference waveform. We
apply a distance prior which is uniform in comoving volume and source frame time
instead of a power law distribution used in \citet{GWTC-2.1, GWTC-3,
GWTC-2.1_dataset, GWTC-3_dataset}, and set the prior on $\kappa$ to a uniform
distribution between $-1$ and $1$.  \wf{Actually, the LIGO references
\emph{also} used the ``cosmological'' prior that is uniform merger rate density
in the source frame; maybe just drop the ``instead of\ldots''?}

@@

Since birefringence is a property of spacetime and we are targeting a
propagation effect not intrinsic to any source, we should expect $\kappa$ to
take the same value for all signals, whether it vanishes or not. \mi{caveats?}
\wf{\emph{If} birefringence were a property of spacetime, this would be true;
but in (some? most? all?) ``parity violating'' theories, the parity violation is
due to an extra \emph{field} that is parity-odd and couples to gravity.  In this
case whether the birefringence is the same along the different lines of sight to
the different events depends on \emph{whatever that field is doing between us
and the events}.  The \emph{simplest} assumption is that the field is ``the
same'' or shared from event-to-event, but it's by no means required that it be
so.  Hence the analysis of the next section.} Under this assumption, the
posterior on $\kappa$ inferred collectively from all events with a uniform prior
is simply obtained from the product of the individual likelihoods, $p(d_i \mid
\kappa) \propto p(\kappa \mid d_i)/p(\kappa)$, such that
\begin{equation}
    p(\kappa \mid \{d_i\})\propto p(\kappa) \prod_{i}\frac{p(\kappa \mid d_i)}{p(\kappa)}\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$\textsuperscript{th} event, and $p(\kappa)$ is the prior on $\kappa$; since the prior is uniform, in our case Eq.~\eqref{eq:restricted_posterior} reduces to the product of the posteriors, namely $p(\kappa \mid \{d_i\}) \propto \prod_{i}p(\kappa \mid d_i)$.
We use Eq.~\eqref{eq:restricted_posterior} to obtain the primary constraint presented in this work.

@@

% Bayesian Hierarchical Modeling
In order to further characterize our set of measurements, and identify potential outliers, we carry out an additional collective analysis that allows $\kappa$ to take different values for each event \cite{Zimmerman:2019wzo,Isi:2022cii}.
To do this, we apply hierarchical Bayesian inference \cite{Loredo:2004nn} to model the distribution of $\kappa$'s consistent with the observed data:
we posit that there is a specific value of the parameter, $\kappa_i$, associated with each event, drawn from some unknown distribution of underlying values; from the imperfect measurements of $\kappa_i$ for each event, we may reconstruct the underlying distribution.
If we are interested in constraining the first two moments of the distribution, it is convenient to parametrize the $\kappa_i$'s as drawn from a Gaussian with unknown mean $\mu$ and variance $\sigma^2$, i.e., $\kappa_i \sim \mathcal{N}(\mu, \sigma^2)$ \cite{Isi:2019asy}, and measure those hyperparameters from the collection of observed data.

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements is statistically unlikely to originate from a unique $\kappa$ value.
This could signal further physics than is implied by Eq.~\eqref{eq:waveform_modification} \mi{which physics? @Will, any ideas?} \wf{see comment above} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

@@

\begin{table}
    \caption{Events that best constrain $\kappa$, sorted by posterior standard deviation $\sigma_i$. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/best_events_kappa.txt}
    \end{ruledtabular}
    \label{tab:best_events_kappa}
\end{table}

@@

\begin{table}
    \caption{Events that show bimodality in the $\kappa$ posterior tend to have high total masses. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}

@@

% Constraint on $\kappa$
Finally, the hierarchical result in Fig.~\ref{fig:corner_Gaussian} can be translated into an overall expectation for $\kappa_i$ via Eq.~\eqref{eq:generic_posterior}.
We show the result of doing this in Fig.~\ref{fig:posterior_kappa}, where we also compare to the posterior on $\kappa$ obtained by assuming a shared value across events (same result shown as a blue band in Fig.~\ref{fig:violin_kappa}).
The hierarchical measurement leads to an expectation that $\kappa_i = $\variable{output/generic_kappa_median.txt}, whereas the shared-$\kappa$ measurement implies $\kappa = $\variable{output/restricted_kappa_median.txt}, both symmetric 90\%-credible intervals around the median.  \wf{It is very interesting to me that the analysis with $\mu$ and $\sigma$ provides a \emph{tighter} constraint on $\kappa$ than the one with the shared-$\kappa$ value.  @Max: maybe we should polish off that note with Katerina and Cole about the benefts / costs of doing the $\mu$-$\sigma$ analysis?  Do either of you understand how this comes about?  (I can make up some b.s.\ about the bimodal posteriors yielding a better model fit for the $\mu$-$\sigma$ analysis than the fixed-$\kappa$ analysis, but I'm not sure it will hold water.)}

@@

It is not clear what specific feature of the GW170818 data leads to this posterior structure.
However, the reason may be related to the fact that this event provides an uncommonly definite measurement of the polarization angle and spin orientations \cite{Varma:2021csh}.
Indeed, the inferred polarization angle and spins in the birefringent case are are notably different from the \ac{GR} case (Fig. XXX).
If our inference on $\kappa$ is indeed tied to those parameters, we might be susceptible to systematics in the modeling of spin angles in \textsc{IMRPhenomXPHM} and might thus benefit from further analysis with other waveforms like \textsc{NRSur7dq4} \cite{Varma:2018mmi}.
In any case, as we argued in Sec.~\ref{sec:results:hier}, there is no evidence at the moment that this event is inconsistent with the either collection of measurements or with $\kappa = 0$ (Fig.~\ref{fig:reweighted_kappa}).

>>>>

We carry out \ac{PE} using a custom version of \textsc{Bilby} \citep{Bilby},
that applies Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline
waveform. We take the \ac{PE} configuration as in \citet{GWTC-2.1, GWTC-3,
GWTC-2.1_dataset, GWTC-3_dataset} as a starting point, with
\textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb} as the reference waveform. We
apply a distance prior which is uniform in comoving volume and source frame time, and set the prior on $\kappa$ to a uniform
distribution between $-1$ and $1$.

@@

In the most simplified scenarios, birefringence is a property of spacetime that is not intrinsic to any source or region in space. 
If so, we should expect $\kappa$ to
take the same value for all signals, whether it vanishes or not.
Under this assumption, the
posterior on $\kappa$ inferred collectively from all events 
is simply obtained from the product of the individual likelihoods, $p(d_i \mid
\kappa) \propto p(\kappa \mid d_i)/p(\kappa)$, such that
\begin{equation}
    p(\kappa \mid \{d_i\})\propto p(\kappa) \prod_{i}\frac{p(\kappa \mid d_i)}{p(\kappa)}\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$\textsuperscript{th} event, and $p(\kappa)$ is the prior on $\kappa$; since the prior is uniform, in our case Eq.~\eqref{eq:restricted_posterior} reduces to the product of the posteriors, namely $p(\kappa \mid \{d_i\}) \propto \prod_{i}p(\kappa \mid d_i)$.
We use Eq.~\eqref{eq:restricted_posterior} to obtain the primary constraint presented in this work.

@@

% Bayesian Hierarchical Modeling
Under many frameworks, birefringence is brokered by an extra parity-odd
field that couples to gravity. In this case, whether the birefringence
is the same along the different lines of sight to the different events depends
on cosmic history and the local evolution of the field. The \emph{simplest}
assumption is that the field manifests equally for all events, as assumed in the previous subsection, but this is by
no means required.
This motivates a collective analysis that does not assume $\kappa$ is shared across events \cite{Zimmerman:2019wzo,Isi:2022cii}.
Such an analysis has the additional advantage of helping us further characterize our set of measurements, and identify potential outliers.

To do this, we apply hierarchical Bayesian inference \cite{Loredo:2004nn} to model the distribution of $\kappa$'s consistent with the observed data:
we posit that there is a specific value of the parameter, $\kappa_i$, associated with each event, drawn from some unknown distribution of underlying values; from the imperfect measurements of $\kappa_i$ for each event, we may reconstruct the underlying distribution.
If we are interested in constraining the first two moments of the distribution, it is convenient to parametrize the $\kappa_i$'s as drawn from a Gaussian with unknown mean $\mu$ and variance $\sigma^2$, i.e., $\kappa_i \sim \mathcal{N}(\mu, \sigma^2)$ \cite{Isi:2019asy}, and measure those hyperparameters from the collection of observed data.

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements is statistically unlikely to originate from a unique $\kappa$ value.
This could signal richer physics than is implied by Eq.~\eqref{eq:waveform_modification} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

@@

\begin{table}[h]
    \caption{Events that best constrain $\kappa$, sorted by posterior standard deviation $\sigma_i$. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/best_events_kappa.txt}
    \end{ruledtabular}
    \label{tab:best_events_kappa}
\end{table}

@@

\begin{table}[h]
    \caption{Events that show bimodality in the $\kappa$ posterior tend to have high total masses. CL is the credible level of $\kappa = 0$.}
    \begin{ruledtabular}
        \variable{output/bimodal_events_mass.txt}
    \end{ruledtabular}
    \label{tab:bimodal_events_mass}
\end{table}

@@

% Constraint on $\kappa$
Finally, the hierarchical result in Fig.~\ref{fig:corner_Gaussian} can be translated into an overall expectation for $\kappa_i$ via Eq.~\eqref{eq:generic_posterior}.
We show the result of doing this in Fig.~\ref{fig:posterior_kappa}, where we also compare to the posterior on $\kappa$ obtained by assuming a shared value across events (same result shown as a blue band in Fig.~\ref{fig:violin_kappa}).
The hierarchical measurement leads to an expectation that $\kappa_i = $\variable{output/generic_kappa_median.txt}, whereas the shared-$\kappa$ measurement implies $\kappa = $\variable{output/restricted_kappa_median.txt}, both symmetric 90\%-credible intervals around the median.

@@

It is not clear what specific feature of the GW170818 data leads to this posterior structure.
However, the reason may be related to the fact that this event provides an uncommonly definite measurement of the polarization angle and spin orientations \cite{Varma:2021csh}.
Indeed, the inferred polarization angle and spins in the birefringent case are notably different from the \ac{GR} case (Fig.~\ref{fig:corner_GW170818_appendix}).
If our inference on $\kappa$ is indeed tied to those parameters, we might be susceptible to systematics in the modeling of spin angles in \textsc{IMRPhenomXPHM} and might thus benefit from further analysis with other waveforms like \textsc{NRSur7dq4} \cite{Varma:2018mmi}.
In any case, as we argued in Sec.~\ref{sec:results:hier}, there is no evidence at the moment that this event is inconsistent with the either collection of measurements or with $\kappa = 0$ (Fig.~\ref{fig:reweighted_kappa}).
